SDSS-IV MaNGA: The Radial Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2020-10-22 maintex.tex
major revision to Sec 4.1.1

In the first method, we only control for the stellar mass between the pairs and controls. We split both the pair and control samples into five evenly spaced stellar mass bins over the range, \logm\ $=$ 9.0$-$11.5. The pairs are then compared to their respective controls within each stellar mass bins. While the redshift is not constrained in this method, we will show in Section \ref{sec:mass-bin} that this method is sufficient to reveal the merger induced star formation.

In the second method, we wish to account for both stellar mass and redshift. To do this we select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within a 0.1~dex stellar mass limit and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. Since the MaNGA sample has a tight distribution in stellar mass and redshift, the stellar mass and MaNGA subsample restriction effectively sets a $\Delta z$ $\lesssim$ $\sim$0.025 redshift tolerance.

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we randomly down-select the total number of acquired control galaxies to 20. A given control galaxy may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 4$-$6 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. Only 5 of the paired galaxies required an extra iteration to find 20 controls.

@@

We then take the difference between the stacked profiles of the paired galaxies and the control galaxies, pair - control, in log space (this means that the difference profiles really represent a ratio between the pairs and controls in linear space). This gives us difference profile, $\Delta$log(sSFR), which shows us where the sSFR is enhanced or suppressed (shown in the right hand panel of Figure \ref{fig:ssfr_prof}). Across all stellar mass bins, the sSFR of paired galaxies are centrally enhanced by $\sim$0.35$\pm$0.1~dex, which gradually falls to zero around $\sim$1.2 \reff. In the outskirts of the pairs beyond 1.2 \reff, the pairs feature lightly suppressed sSFR of 0.0$-$0.1 dex. All of the mass ranges except for the highest mass range, \logm\ $=$ 11.0$-$11.5, trend closely to the median profile. For the most massive galaxies, \logm\ $=$ 11.0$-$11.5, the enhancement to the sSFR within 0.5 \reff\ is higher than the median profile at 0.5$-$0.6 dex. The sSFR enhancement in the sample seems to be mass independent except for the most massive galaxies which feature higher levels of sSFR enhancement.

Alternatively, the central enhancement of the sSFR in the paired galaxies can be interpreted as the quenching of star formation in the centers of isolated galaxies. While the mechanism for quenching in galaxies remains under debate, the merger process seems to be capable of rejuvenating star formation in the centers of these galaxies, resulting in the flatter sSFR profiles.

The higher rate of star formation within interacting galaxies will mean that the stellar mass in the interacting galaxies will grow at a faster rate compared to secular mass growth. The fractional mass growth rate is simply the difference between the sSFR profiles of the paired and control galaxies in linear space (pairs - controls) multiplied by a given timescale. We chose a timescale of 0.4 Gyr which is estimated to be the average timescale for the merger induced starburst \citep{Feng:2019}.

Since the $\Delta$log(sSFR) is independent of galaxy mass, the fractional mass change will depend on the stellar mass of the paired galaxy. That is, since the amount of new stellar mass from merger induced interactions is roughly constant across galaxies of different stellar masses, lower mass galaxies will experience a greater change to their total stellar mass. Low mass galaxies, \logm\ $=$ 9.0$-$9.5, experience a fractional mass change of $\Delta M_{\rm merger}$/$M_{\rm total}$ $=$ 0.03 (3\%) while massive galaxies, \logm\ $=$ 11.0$-$11.5, experience a fractional mass change of $\Delta M_{\rm merger}$/$M_{\rm total}$ $=$ 0.001 (0.1\%). This shows that the merger induced mass growth is greatest in low mass galaxies. Further, the merger induced mass growth is restricted to the centers of the paired galaxies which will result in galaxies which have more prominent bulges in their centers.

This mass-binning method has the benefit of a high S/N control profile since each mass bin will have hundreds of control galaxies while the median control profiles of the second method will only have 20 controls. On the other hand, this method only controls for a single galaxy parameter, stellar mass, though it is still relevant that the central enhancement of the sSFR in pairs is able to be observed. The profiles can also only be studied as a function of stellar mass so we are unable to study the profiles as a function of projected separation or mass ratio.

>>>>

In the first method, we simply control for the stellar mass between the pairs and controls. We split both the pair and control samples into five evenly spaced stellar mass bins over the range, \logm\ $=$ 9.0$-$11.5. The pairs are then compared to their respective controls within each stellar mass bins. While the redshift is not constrained in this method, we will show in Section \ref{sec:mass-bin} that this method is sufficient to reveal the merger induced star formation.

In the second method, we select a ``tailored'' control sample for each paired galaxy from the full sample of isolated galaxies. We select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within a 0.1~dex stellar mass limit and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. Since the MaNGA sample has a tight distribution in stellar mass and redshift, the stellar mass and MaNGA subsample restriction effectively sets a $\Delta z$ $\lesssim$ $\sim$0.025 redshift tolerance. With these requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we randomly down-select the total number of acquired control galaxies to 20. A given control galaxy may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 4$-$6 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. Only 5 of the paired galaxies required an extra iteration to find 20 controls.

@@

We then take the difference between the stacked profiles of the paired galaxies and the control galaxies, pair - control, in log space (this means that the difference profiles really represent a ratio between the pairs and controls in linear space). This gives us difference profile, $\Delta$log(sSFR), which shows us where the sSFR is enhanced or suppressed (shown in the right hand panel of Figure \ref{fig:ssfr_prof}). Across all stellar mass bins, the sSFR of paired galaxies are centrally enhanced by $\sim$0.35$\pm$0.1~dex, which gradually falls to zero around $\sim$1.2 \reff. In the outskirts of the pairs beyond 1.2 \reff, the pairs feature lightly suppressed sSFR of 0.0$-$0.1 dex. All of the mass ranges except for the highest mass range, \logm\ $=$ 11.0$-$11.5, trend closely to the median profile. For the most massive galaxies, \logm\ $=$ 11.0$-$11.5, the enhancement to the sSFR within 0.5 \reff\ is significantly higher than the median profile (reaching 0.5$-$0.6 dex). As will be discussed in \S~\ref{sec:tailored}, we believe this elevated enhancement in the highest mass bin is driven by more compact mergers instead of stellar mass.

In summary, Figure~\ref{fig:ssfr_prof} shows that the merger process is capable of rejuvenating star formation in the centers of these galaxies, resulting in almost flat sSFR radial profiles. In addition, the sSFR enhancement is mass independent -- star formation proceeds at $\sim$2$\times$ (0.3~dex) greater rates in galaxies with close companions than in more isolated galaxies, across a wide mass range between 9.0 $\leq$ \logm\ $\leq$ 11.5.

This method of stacking sSFR profiles by a single parameter (stellar mass) has the advantage of simplicity and large statistical samples. It has revealed the first order result of mass-independent, centrally enhanced star formation in close galaxy pairs. To proceed with exploring the dependency of $\Delta$sSFR on merger parameters such as separation and mass ratio, we will utilize a more sophisticated method of selecting control samples for individual paired galaxies based on stellar mass and redshift in the following subsections.
